\bf Mechanistic models for panel data: analysis of ecological experiments with four interacting species

### 2024-04-11 ms.Rnw
Update of paper

<<Setup, include=FALSE,echo=FALSE,results=hide>>=
 library(plyr)
 library(reshape2)
library(patchwork)
 library(magrittr)
 library(foreach)
 library(readxl)
 library(doParallel)
 registerDoParallel(cores=36)
 library(pomp)
 library(panelPomp)
 library(ggplot2)
 library(gridExtra)

@@

As juveniles rarely exhibit clear signs of infection, only adult \textit{Daphnia} were used for quantifying effects in both Non-parasitized and Parasitized Dynamics to minimize information bias. Figures below illustrates these dynamics, with the figure 1 depicting Non-parasitized Dynamics (mixed-species) where changes in native and invasive host densities reflect direct resource competition, for example, food alga \textit{Ankistrodesmus falcatus}. Figure 2 shows Parasitized Dynamics, where the interplay of susceptible and infected \textit{Daphnia} of different species is further influenced by parasites, complicating the variation of susceptible and infected \textit{Daphnia} of different races, which makes it difficult to directly analyze the sources of influence on their variation. Non-parasitized Dynamics will be analyzed at the first place to help understand the effect of resource competition in mixed-species. And based on this result, the correlation between native host and invasive host in Parasitized Dynamics will be further explored.

\begin{figure}[H]
\centering
<<echo=FALSE, fig.cap="Plot1", fig.align='center', fig=TRUE, width=16, height=10>>=  
   mixed_nopara_data_combined <- do.call(rbind, lapply(1:length(mixed_nopara_data), function(i) {
     cbind(mixed_nopara_data[[i]], group=paste("Trial", trails[i]))
   }))
    mixed_para_data_combined <- do.call(rbind, lapply(1:length(mixed_para_data), function(i) {
     cbind(mixed_para_data[[i]], group=paste("Trial", trails[i]))
   }))

p1<- ggplot(mixed_nopara_data_combined, aes(x=day)) + 
   geom_line(aes(y=dent.adult, group=interaction(group, "dent.adult"))) + 
   geom_line(aes(y=lum.adult, color="blue", group=interaction(group, "lum.adult"))) +
   scale_x_continuous(limits = c(0, 52)) +
   scale_y_continuous(limits = c(0, 175)) +
   facet_grid(. ~ group, scales = "fixed", space = "free") +
   theme(
     strip.text = element_text(size=10),
     axis.text.x=element_text(color="black"),
     axis.ticks.x=element_line(color="black"),
     axis.text.y=element_text(color="black"),
     axis.ticks.y=element_line(color="black"),
     axis.title.x=element_blank(),
     axis.title.y=element_blank(),
     legend.position="none",
     panel.spacing = unit(0, "lines"), # Adjust spacing between panels
     strip.background = element_blank(), # Optional: Remove facet label background
     strip.placement = "outside"
   ) +
   labs(title="Daphnia density in Non-parasitized Dynamics") +
   guides(colour=guide_legend(title="Species"))

p2<- ggplot(mixed_para_data_combined, aes(x=day)) + 
   geom_line(aes(y=dentadult, group=interaction(group, "dentadult"))) + 
   geom_line(aes(y=lumadult, color="blue", group=interaction(group, "lumadult"))) +
    geom_line(aes(y=luminf,color="green", group=interaction(group, "luminf"))) + 
   geom_line(aes(y=dentinf, color="red", group=interaction(group, "dentinf"))) +

   scale_x_continuous(limits = c(0, 52)) +
   scale_y_continuous(limits = c(0, 175)) +
   facet_grid(. ~ group, scales = "fixed", space = "free") +
   theme(
     strip.text = element_text(size=10),
     axis.text.x=element_text(color="black"),
     axis.ticks.x=element_line(color="black"),
     axis.text.y=element_text(color="black"),
     axis.ticks.y=element_line(color="black"),
     axis.title.x=element_blank(),
     axis.title.y=element_blank(),
     legend.position="none",
     panel.spacing = unit(0, "lines"), # Adjust spacing between panels
     strip.background = element_blank(), # Optional: Remove facet label background
     strip.placement = "outside"
   ) +
   labs(title="Daphnia density in Parasitized Dynamics") +
   guides(colour=guide_legend(title="Species"))
grid.arrange(p1, p2, ncol = 1)
@
\caption{Plot of density of hosts in dynamics including both native and invasive host species of \textit{Daphnia}. The trail A to J represents different buckets in different experiment. The black lines represent the \textit{D. dentifera} density, the red lines show the \textit{D. lumholtzi} density, the green lines are the infected \textit{D. dentifera} density, and the blue lines indicate the infected \textit{D. lumholtzi} density.}
\end{figure}

@@

To analyze the dynamics of Parasitized and Non-parasitized systems, the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model, an adaptation of the SIR model \citep{weiss2013sir}, is utilized. This model integrates the aspects of food and parasites, unlike the classical SIR model, to more accurately reflect the interactions between native and invasive hosts in the context of resource availability and parasitic influence. The SIRPF model incorporates these factors as quantitative latent state variables, enabling a more comprehensive understanding of the system's dynamics. In the meantime, the ecological processes of \textit{Daphnia} can be viewed as a Markov process, where the population of Susceptible and Infected individuals is observable, while the amount of food and parasites represent unobservable latent states. Utilizing panel data on \textit{Daphnia}, the PanelPOMP framework becomes a valuable analytical tool \citep{Carles2020}. Despite the lack of direct measurement data for these latent states, PanelPOMP, which has proven efficiency in analyzing ecological panel data, facilitates a robust analysis of these dynamics, effectively handling the complexities inherent in ecological systems.\\

The SIRPF model describes the dynamics of the Susceptible($S$), Infected($I$), Recovered, Food($F$), and Parasite($P$) populations as the following graph and model show:

@@

{
\small
\begin{align}
&d S(t) = r F(t)\left(S(t)+\xi I(t)\right)dt - \theta_{S}S(t) dt - p f_{S} S(t) Pdt - \delta S(t)dt + S(t) d\zeta_{S},\\
&d I(t) = p f_{S} S(t) Pdt - \theta_{I} I(t)dt - \delta I(t)dt + I(t) d\zeta_{I},\\
&dF(t) = \alpha F(t)\left(1-\frac{F(t)}{K_F}\right)dt - \gamma F(t) \left(S(t)+\xi I(t)\right) dt - \delta F(t) dt + \mu dt + F(t) d\zeta_F,\\
&d P(t) = \beta \theta_I I(t)dt - f_{S} S(t) P(t)dt - f_{I} I(t) P(t)dt - \theta_{\mathrm{p}} P(t)dt - \delta P(t) dt + P(t) d\zeta_{P},\\
&d\zeta_{S} \sim N\left(0, \sigma_{S}^{2}dt\right),\\
&d\zeta_{I} \sim N\left(0, \sigma_{I}^{2}dt\right),\\
&d\zeta_{F} \sim N\left(0, \sigma_{F}^{2}dt\right),\\
&d\zeta_{P} \sim N\left(0, \sigma_{P}^{2}dt\right).
\end{align}
}

Each equation intricately ties the state variables to depict the complex dynamics of the single species Parasitized system. Building on the SIRPF model, the SIRPF2 model further adapts to a mixed-race Parasitized system by introducing another \textit{Daphnia} species. This advanced model incorporates both invasive($S_i , I_i$) and native($S_n , I_n$)hosts, along with food($F$) and parasites($P$), and delineates their specific interactions within the ecosystem as the following graph and model shows. In the SIRPF2 model, the subscript n is used to denote parameters and variables related to the native host, i for those pertaining to the invasive host, and p for the parasite spores. The SIRPF2 model thus offers a more detailed representation of the ecological dynamics by accounting for the complex interplay between different species and environmental factors in a mechanistic framework.
\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.5cm, minimum height=0.5cm, rounded corners=.1cm, fill=blue!8},
  rhombus/.style={diamond, draw=black, minimum width=0.1cm, minimum height=0.1cm, fill=purple!8,aspect = 1},
  travel/.style={circle, draw=black, minimum width=0.5cm, minimum height=0.5cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.6cm,inner sep=0cm},
  bendy/.style={bend left=10},
  bendy2/.style={bend left=100},
  bendy3/.style={bend left=-100},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF2 model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. The flow diagram shows the how two species of \textit{Daphnia} interact with each other indirectly through $F$ and $P$. The marginal processes have been shown in SIRPF mode, and SIRPF2 model is the adaption form in mixed-species dynamcis}
\label{fig:flow_SIRPF2}
\end{figure}

{
\small
\begin{align}
% \begin{split}
&d S_n(t) = r_n F(t) \left(S_n(t)+\xi I_n(t)\right)dt - \theta_{S_n}S_n(t) dt - p_n f_{S_n} S_n(t) P(t)dt -\delta S_n(t)dt + S_n(t) d\zeta_{S_n},\\
% \end{split}
% \\
&d I_n(t) = p_n f_{S_n} S_n(t) Pdt - \theta_{I_n} I_n(t)dt - \delta I_n(t)dt + I_n d\zeta_{I_n},\\
% \begin{split}
&d S_i(t) = r_i F(t)\left(S_i(t)+\xi I_i(t)\right)dt - \theta_{S_i} S_i(t) dt - p_i f_{S_i} S_i(t) P(t)dt -  \delta S_i(t)dt + S_i(t) d\zeta_{S_i},\\
% \end{split}
% \\
&d I_i(t) = p_i f_{S_i} S_i(t) P(t)dt - \theta_{I_i} I_i(t)dt - \delta I_i(t)dt + I_i(t) d\zeta_{I_i}(t),\\
\begin{split}
&dF(t) = \alpha F(t)\left(1-\frac{F(t)}{K_F}\right)dt - \gamma_n F(t) \left(S_n(t)+\xi I_n(t)\right) dt - \gamma_i F(t) \left(S_i(t)+\xi I_i(t)\right) dt\\
&\quad \quad \quad- \delta F(t) dt + \mu dt + F(t) d\zeta_F,\\
\end{split}
\\
\begin{split}
&d P(t) = \beta_n \theta_{I_n} I_n(t)dt + \beta_i \theta_{I_i} I_i(t)dt - f_{S_n} S_n(t) P(t)dt - f_{I_n} I_n(t) P(t)dt - f_{S_i} S_i(t) P(t)dt \\
&\quad \quad \quad- f_{\mathrm{I}_i} I_i(t) P(t)dt - \theta_{\mathrm{p}} P(t)dt - \delta P(t) dt + P(t) d\zeta_{P},\\
\end{split}
\\
&d\zeta_{S_n} \sim N\left(0, \sigma_{S_n}^{2}dt\right),\\
&d\zeta_{I_n} \sim N\left(0, \sigma_{I_n}^{2}dt\right),\\
&d\zeta_{S_i} \sim N\left(0, \sigma_{S_i}^{2}dt\right),\\
&d\zeta_{I_i} \sim N\left(0, \sigma_{I_i}^{2}dt\right),\\
&d\zeta_{F} \sim N\left(0, \sigma_{F}^{2}dt\right),\\
&d\zeta_{P} \sim N\left(0, \sigma_{P}^{2}dt\right).
\end{align}
}

@@

<<Results, include=FALSE,echo=FALSE,results=hide>>=
load('all_shared.RData')

@@

parms$K_F = parms$K_F * 1000
parms$ri = parms$ri * 10
parms$rn = parms$rn * 10
parms$theta_Ii = parms$theta_Ii * 10
parms$theta_In = parms$theta_In * 10
parms$theta_Si = parms$theta_Si * 10
parms$theta_Sn = parms$theta_Sn * 10
parms$theta_P = parms$theta_P * 1000
parms$probi = parms$probi * 10
parms$probn = parms$probi * 10
parms$f_Sn = parms$f_Sn * 10000
parms$f_Si = parms$f_Si * 10000
parms$f_In = parms$f_In * 1000
parms$f_Ii = parms$f_Ii * 1000000
parms$alpha = parms$alpha * 10000
parms$gamman = parms$gamman * 1e8
parms$gammai = parms$gammai * 1e4

parms$sigSi = parms$sigSi * 1e4
parms$sigIn = parms$sigIn * 1e5
parms$sigIi = parms$sigIi * 1e5
parms$sigF = parms$sigF * 10
parms$sigP = parms$sigP * 100
@

@@

$S_j$ & Susceptible host density for species $j$  &$individuals \cdot L ^{-1}$ & Variable\\
$I_j$ & Infected host density for species $j$ & $ individuals \cdot L ^{-1}$& Variable\\
$F$& Alga density& $10^5 cells \cdot L ^{-1}$& Variable\\
$P$ & Spore density & $10^3 spores \cdot L ^{-1}$& Variable\\
$K_F$ & Alga carrying capacity&$10^2\cdot cells \cdot L ^{-1} $ & \Sexpr{mysignif(parms$K_F,2)}\\
$r_n$ & Native growth rate &$10^{-6} \cdot cells^{-1} \cdot day^{-1}$ & \Sexpr{mysignif(parms$rn,2)}\\
$r_i$ & Invasive growth rate &$10^{-6} \cdot cells^{-1} \cdot day^{-1}$ & \Sexpr{mysignif(parms$ri,2)}\\
$\theta_{I_n}$ & Native infected host mortality rate&$ 10^{-1} \cdot day^{-1}$  & \Sexpr{mysignif(parms$theta_In,2)}\\
$\theta_{I_i}$ & Invasive infected host mortality rate&$10^{-1} \cdot day^{-1}$  & \Sexpr{mysignif(parms$theta_Ii,2)}\\
$\theta_{S_i}$ & Invasive susceptible host mortality rate&$10^{-1} \cdot day^{-1}$ & \Sexpr{mysignif(parms$theta_Si,2)}\\
$\theta_{S_n}$ & Native susceptible host mortality rate&$10^{-1} \cdot day^{-1}$ & \Sexpr{mysignif(parms$theta_Sn,2)}\\
$\theta_{P}$ & Spore degradation rate&$10^{-1} \cdot  day^{-3}$ & \Sexpr{mysignif(parms$theta_P,2)}\\
$p_n$ &Native probability of infection per spore &$10^{-4} \cdot individuals \cdot spore^{-1} $ & \Sexpr{mysignif(parms$probn,2)}\\
$p_i$ & Invasive probability of infection per spore&$10^{-4} \cdot individuals \cdot spore^{-1} $ & \Sexpr{mysignif(parms$probi,2)}\\
$f_{S_n}$ & Native susceptible host filtering rate&$10^{-4} \cdot L \cdot day^{-1} \cdot individuals^{-1} $ & \Sexpr{mysignif(parms$f_Sn,2)}\\
$f_{S_i}$ & Invasive susceptible host filtering rate&$10^{-4} \cdot L \cdot day^{-1} \cdot individuals^{-1} $& \Sexpr{mysignif(parms$f_Si,2)} \\
$f_{I_n}$ & Native infected host filtering rate&$10^{-3} \cdot L \cdot day^{-1} \cdot individuals^{-1} $& \Sexpr{mysignif(parms$f_In,2)} \\
$f_{I_i}$ & Invasive infected host filtering rate&$10^{-6} \cdot L \cdot day^{-1} \cdot individuals^{-1} $ & \Sexpr{mysignif(parms$f_Ii,2)}\\

@@

$\alpha$ & Alga maximum exponential growth rate&$10^{-4} \cdot day ^{-1} $& \Sexpr{mysignif(parms$alpha,2)} \\
$\gamma_n$ & Native host consumption rate&$10^{-4} \cdot L \cdot individual^{-1} \cdot day ^{-1} $ & \Sexpr{mysignif(parms$gamman,2)}\\
$\gamma_i$ & Invsaive host consumption rate&$10^{-4} \cdot L \cdot individual^{-1} \cdot day ^{-1} $ & \Sexpr{mysignif(parms$gammai,2)}\\
$\mu$ & Alag refilling rate&$10^5 cells \cdot L ^{-1} \cdot day^{-1}$ & 0.37\\
$\xi$ &Reduction in infected host reproduction and consumption& $Unitless$ & \Sexpr{mysignif(parms$xi,2)}\\
$\beta_i$ & Spores produced per infected invasive species&$10^3 \cdot spores \cdot individuals^{-1} $ & \Sexpr{mysignif(parms$betai,2)}\\
$\beta_n$ & Spores produced per infected native species&$10^3 \cdot spores \cdot individuals^{-1} $ & \Sexpr{mysignif(parms$betan,2)}\\
$\sigma_{S_n}$ & Standard deviation of Brownian motion of susceptible native&$\sqrt{individual \cdot t^{-1}}$ & \Sexpr{mysignif(parms$sigSn,2)}\\
$\sigma_{S_i}$ & Standard deviation of Brownian motion of susceptible invasive &$10^{-4} \cdot \sqrt{individual \cdot t^{-1}}$ & \Sexpr{mysignif(parms$sigSi,2)}\\

>>>>

<<Setup, include=FALSE,echo=FALSE,results=hide>>=
 library(plyr)
 library(reshape2)
library(patchwork)
 library(magrittr)
 library(foreach)
 library(readxl)
 library(doParallel)
 registerDoParallel(cores=36)
 library(pomp)
 library(panelPomp)
 library(ggplot2)
 library(gridExtra)
 library(scales)  # For formatting scales
 library(patchwork)

@@

As juveniles rarely exhibit clear signs of infection, only adult \textit{Daphnia} were used for quantifying effects in both non-parasitized and prarasitized dynamics to minimize information bias. Figures below illustrates these dynamics, with the first row of Figure 1 depicting non-parasitized dynamics (mixed-species) where changes in native and invasive host densities reflect direct resource competition, for example, food alga \textit{Ankistrodesmus falcatus}. The second row of Figure 1 shows parasitized dynamics, where the interplay of susceptible and infected \textit{Daphnia} of different species is further influenced by parasites, complicating the variation of susceptible and infected \textit{Daphnia} of different races, which makes it difficult to directly analyze the sources of influence on their variation. Non-parasitized dynamics will be analyzed at the first place to help understand the effect of resource competition in mixed-species. And based on this result, the correlation between native host and invasive host in parasitized dynamics will be further explored.

\begin{figure}[H]
\centering
<<echo=FALSE, fig.cap="Plot1", fig=TRUE, width=25, height=10,out.width='\\textwidth'>>=
mixed_nopara_data_combined <- do.call(rbind, lapply(1:length(mixed_nopara_data), function(i) {
  cbind(mixed_nopara_data[[i]], Trial = paste("Trial", trails[i]))
}))
mixed_nopara_data_combined$Bucket = paste0(rep("Bucket ",80),rep(1:9, each = 10))

mixed_para_data_combined <- do.call(rbind, lapply(1:length(mixed_para_data), function(i) {
  cbind(mixed_para_data[[i]], Trial = paste("Trial", trails[i]))
}))

mixed_para_data_combined$Bucket = paste0(rep("Bucket ",80),rep(1:8, each = 10))

bucket_labeller <- as_labeller(function(bucket) gsub("Bucket ", "Bucket-", bucket))
# Adjusting p1 with specified modifications
library(ggplot2)

# Adjusting p1 with specified font size modifications
p1 <- ggplot(mixed_nopara_data_combined, aes(x = day, y = sqrt(dent.adult))) +
  geom_line(aes(color = "Dent Adult"), size = 1) +
  geom_line(aes(y = sqrt(lum.adult), color = "Lum Adult"), size = 1) +
  facet_wrap(~Bucket, nrow = 1, labeller = bucket_labeller) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 10),  # Y-axis title
        axis.text.x = element_text(size = 10),   # X-axis text
        axis.text.y = element_text(size = 10),   # Y-axis text
        legend.title = element_text(size = 10),  # Legend title
        legend.text = element_text(size = 10),   # Legend text
        strip.text = element_text(size = 10),    # Facet label text
        strip.background = element_blank(),
        strip.placement = "outside") +
  ylab('Square root of Daphnia density') +
  scale_x_continuous(limits = c(0, 52), breaks = c(0, 25, 50)) +
  labs(title = "Daphnia Density in Non-parasitized Dynamics", color = "Species") +
  scale_color_manual(values = c("Dent Adult" = "black", "Lum Adult" = "blue")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Species"))

# Adjusting p2 with specified font size modifications
p2 <- ggplot(mixed_para_data_combined, aes(x = day, y = sqrt(dentadult))) +
  geom_line(aes(color = "Dent Adult"), size = 1) +
  geom_line(aes(y = sqrt(lumadult), color = "Lum Adult"), size = 1) +
  geom_line(aes(y = sqrt(luminf), color = "Lum Inf"), size = 1) +
  geom_line(aes(y = sqrt(dentinf), color = "Dent Inf"), size = 1) +
  facet_wrap(~Bucket, nrow = 1, labeller = bucket_labeller) +
  theme(axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        strip.text = element_text(size = 10),
        strip.background = element_blank(),
        strip.placement = "outside") +
  ylab('Square root of Daphnia density') +
  scale_x_continuous(limits = c(0, 52), breaks = c(0, 25, 50)) +
  labs(title = "Daphnia Density in Parasitized Dynamics", color = "Species") +
  scale_color_manual(values = c("Dent Adult" = "black", "Lum Adult" = "blue", "Lum Inf" = "green", "Dent Inf" = "red")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Species"))

grid.arrange(p1, p2, ncol = 1)
@
% \includegraphics[width=1\textwidth]{"Density_plot.png"} 
% \label{fig:plot1}
\caption{Plot of density of hosts in dynamics including both native and invasive host species of \textit{Daphnia}. The Buckets 1 to 9 represents different buckets in different experiment. The black lines represent the \textit{D. dentifera} density, the red lines show the \textit{D. lumholtzi} density, the green lines are the infected \textit{D. dentifera} density, and the blue lines indicate the infected \textit{D. lumholtzi} density.}
\end{figure}

@@

To analyze the dynamics of parasitized and non-parasitized systems, the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model, an adaptation of the SIR model \citep{weiss2013sir}, is utilized. This model integrates the aspects of food and parasites, unlike the classical SIR model, to more accurately reflect the interactions between native and invasive hosts in the context of resource availability and parasitic influence. The SIRPF model incorporates these factors as quantitative latent state variables, enabling a more comprehensive understanding of the system's dynamics. In the meantime, the ecological processes of \textit{Daphnia} can be viewed as a Markov process, where the population of Susceptible and Infected individuals is observable, while the amount of food and parasites represent unobservable latent states. Utilizing panel data on \textit{Daphnia}, the PanelPOMP framework becomes a valuable analytic tool \citep{Carles2020}. Despite the lack of direct measurement data for these latent states, PanelPOMP, which has proven efficiency in analyzing ecological panel data, facilitates a robust analysis of these dynamics, effectively handling the complexities inherent in ecological systems.\\

The SIRPF model describes the dynamics of the Susceptible($S$), Infected($I$), Recovered, Food($F$), and Parasite($P$) populations as the Figure 2 and model show:

@@

{
\small
\begin{align}
&d S(t) = r F(t)\left(S(t)+\xi I(t)\right)dt - \theta_{S}S(t) dt - p f_{S} S(t) Pdt - \delta S(t)dt + S(t) d\zeta_{S}\\
&d I(t) = p f_{S} S(t) Pdt - \theta_{I} I(t)dt - \delta I(t)dt + I(t) d\zeta_{I}\\
&dF(t) = \alpha F(t)\left(1-\frac{F(t)}{K_F}\right)dt - \gamma F(t) \left(S(t)+\xi I(t)\right) dt - \delta F(t) dt + \mu dt + F(t) d\zeta_F\\
&d P(t) = \beta \theta_I I(t)dt - f_{S} S(t) P(t)dt - f_{I} I(t) P(t)dt - \theta_{\mathrm{p}} P(t)dt - \delta P(t) dt + P(t) d\zeta_{P}\\
&d\zeta_{S} \sim N\left(0, \sigma_{S}^{2}dt\right),\\
&d\zeta_{I} \sim N\left(0, \sigma_{I}^{2}dt\right),\\
&d\zeta_{F} \sim N\left(0, \sigma_{F}^{2}dt\right),\\
&d\zeta_{P} \sim N\left(0, \sigma_{P}^{2}dt\right).
\end{align}
}

Each equation intricately ties the state variables to depict the complex dynamics of the single species parasitized system. Building on the SIRPF model, the SIRPF2 model further adapts to a mixed-race parasitized system by introducing another \textit{Daphnia} species. This advanced model incorporates both invasive($S_i , I_i$) and native($S_n , I_n$)hosts, along with food($F$) and parasites($P$), and delineates their specific interactions within the ecosystem as the following graph and model shows. In the SIRPF2 model, the subscript n is used to denote parameters and variables related to the native host, i for those pertaining to the invasive host, and p for the parasite spores. The SIRPF2 model thus offers a more detailed representation of the ecological dynamics by accounting for the complex interplay between different species and environmental factors in a mechanistic framework.
\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.5cm, minimum height=0.5cm, rounded corners=.1cm, fill=blue!8},
  rhombus/.style={diamond, draw=black, minimum width=0.1cm, minimum height=0.1cm, fill=purple!8,aspect = 1},
  travel/.style={circle, draw=black, minimum width=0.5cm, minimum height=0.5cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.6cm,inner sep=0cm},
  bendy/.style={bend left=10},
  bendy2/.style={bend left=100},
  bendy3/.style={bend left=-100},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF2 model. Apart from the quantified states above, the graph includes $R$ state, which represents death status. The flow diagram shows the how two species of \textit{Daphnia} interact with each other indirectly through $F$ and $P$. The marginal processes have been shown in SIRPF mode, and SIRPF2 model is the adaption form in mixed-species dynamics}
\label{fig:flow_SIRPF2}
\end{figure}

{
\small
\begin{align}
\begin{split}
&d S_n(t) = r_n F(t) \left(S_n(t)+\xi I_n(t)\right)dt - \theta_{S_n}S_n(t) dt - p_n f_{S_n} S_n(t) P(t)dt -\delta S_n(t)dt \\
& \quad \quad \quad + S_n(t) d\zeta_{S_n}\\
\end{split}\\
% \\
&d I_n(t) = p_n f_{S_n} S_n(t) Pdt - \theta_{I_n} I_n(t)dt - \delta I_n(t)dt + I_n d\zeta_{I_n}\\
% \begin{split}
&d S_i(t) = r_i F(t)\left(S_i(t)+\xi I_i(t)\right)dt - \theta_{S_i} S_i(t) dt - p_i f_{S_i} S_i(t) P(t)dt -  \delta S_i(t)dt + S_i(t) d\zeta_{S_i}\\
% \end{split}
% \\
&d I_i(t) = p_i f_{S_i} S_i(t) P(t)dt - \theta_{I_i} I_i(t)dt - \delta I_i(t)dt + I_i(t) d\zeta_{I_i}(t)\\
\begin{split}
&dF(t) = \alpha F(t)\left(1-\frac{F(t)}{K_F}\right)dt - \gamma_n F(t) \left(S_n(t)+\xi I_n(t)\right) dt - \gamma_i F(t) \left(S_i(t)+\xi I_i(t)\right) dt\\
&\quad \quad \quad- \delta F(t) dt + \mu dt + F(t) d\zeta_F\\
\end{split}
\\
\begin{split}
&d P(t) = \beta_n \theta_{I_n} I_n(t)dt + \beta_i \theta_{I_i} I_i(t)dt - f_{S_n} S_n(t) P(t)dt - f_{I_n} I_n(t) P(t)dt - f_{S_i} S_i(t) P(t)dt \\
&\quad \quad \quad- f_{\mathrm{I}_i} I_i(t) P(t)dt - \theta_{\mathrm{p}} P(t)dt - \delta P(t) dt + P(t) d\zeta_{P}\\
\end{split}
\\
&d\zeta_{S_n} \sim N\left(0, \sigma_{S_n}^{2}dt\right),\\
&d\zeta_{I_n} \sim N\left(0, \sigma_{I_n}^{2}dt\right),\\
&d\zeta_{S_i} \sim N\left(0, \sigma_{S_i}^{2}dt\right),\\
&d\zeta_{I_i} \sim N\left(0, \sigma_{I_i}^{2}dt\right),\\
&d\zeta_{F} \sim N\left(0, \sigma_{F}^{2}dt\right),\\
&d\zeta_{P} \sim N\left(0, \sigma_{P}^{2}dt\right).
\end{align}
}

@@

<<Results, include=FALSE,echo=FALSE,results=hide>>=
load('all_shared.RData')
mif.estimate = as.data.frame(t(mif.estimate))

@@

parms$K_F = parms$K_F * 10
parms$ri = parms$ri * 10
parms$rn = parms$rn * 10
parms$theta_Ii = parms$theta_Ii * 10
parms$theta_In = parms$theta_In * 10
parms$theta_Si = parms$theta_Si * 10
parms$theta_Sn = parms$theta_Sn * 10
parms$theta_P = parms$theta_P * 1000
parms$probi = parms$probi * 10
parms$probn = parms$probi * 10
parms$f_Sn = parms$f_Sn * 1e5
parms$f_Si = parms$f_Si * 1e4
parms$f_In = parms$f_In * 1e5
parms$f_Ii = parms$f_Ii * 1e4
parms$alpha = parms$alpha * 1e2
parms$gamman = parms$gamman * 1e7
parms$gammai = parms$gammai * 1e6
parms$sigSn = parms$sigSn * 1e2
parms$sigSi = parms$sigSi * 1e1
parms$sigIn = parms$sigIn * 1e6
parms$sigIi = parms$sigIi * 1e1
parms$sigF = parms$sigF * 1e2
parms$sigP = parms$sigP * 1e6

@

@@

$S_j$ & Susceptible host density for species $j$  &$\mathrm{individual} \cdot L ^{-1}$ & Variable\\
$I_j$ & Infected host density for species $j$ & $ \mathrm{individual} \cdot L ^{-1}$& Variable\\
$F$& Alga density& $10^6 \mathrm{cell} \cdot L ^{-1}$& Variable\\
$P$ & Spore density & $10^3 \mathrm{spore} \cdot L ^{-1}$& Variable\\
$K_F$ & Alga carrying capacity&$10^{-1}\cdot \mathrm{cell} \cdot \mathrm{L} ^{-1} $ & \Sexpr{mysignif(parms$K_F,2)}\\
$r_n$ & Native growth rate &$10^{-7} \cdot  \mathrm{L} \cdot \mathrm{cell}^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$rn,2)}\\
$r_i$ & Invasive growth rate &$10^{-7} \cdot  \mathrm{L} \cdot  \mathrm{cell}^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$ri,2)}\\
$\theta_{I_n}$ & Native infected host mortality rate&$ 10^{-1} \cdot \mathrm{day}^{-1}$  & \Sexpr{mysignif(parms$theta_In,2)}\\
$\theta_{I_i}$ & Invasive infected host mortality rate&$10^{-1} \cdot \mathrm{day}^{-1}$  & \Sexpr{mysignif(parms$theta_Ii,2)}\\
$\theta_{S_i}$ & Invasive susceptible host mortality rate&$10^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$theta_Si,2)}\\
$\theta_{S_n}$ & Native susceptible host mortality rate&$10^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$theta_Sn,2)}\\
$\theta_{P}$ & Spore degradation rate&$10^{-3} \cdot  \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$theta_P,2)}\\
$p_n$ &Native probability of infection per spore &$10^{-4} \cdot \mathrm{individual} \cdot \mathrm{spore}^{-1} $ & \Sexpr{mysignif(parms$probn,2)}\\
$p_i$ & Invasive probability of infection per spore&$10^{-4} \cdot \mathrm{individual} \cdot \mathrm{spore}^{-1} $ & \Sexpr{mysignif(parms$probi,2)}\\
$f_{S_n}$ & Native susceptible host filtering rate&$10^{-5} \cdot \mathrm{L}  \cdot \mathrm{individual}^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$f_Sn,2)}\\
$f_{S_i}$ & Invasive susceptible host filtering rate&$10^{-4} \cdot \mathrm{L} \cdot \mathrm{individual}^{-1} \cdot \mathrm{day}^{-1} $& \Sexpr{mysignif(parms$f_Si,2)} \\
$f_{I_n}$ & Native infected host filtering rate&$10^{-5} \cdot \mathrm{L} \cdot \mathrm{individual}^{-1} \cdot \mathrm{day}^{-1} $& \Sexpr{mysignif(parms$f_In,2)} \\
$f_{I_i}$ & Invasive infected host filtering rate&$10^{-4} \cdot \mathrm{L} \cdot \cdot \mathrm{individual}^{-1} \cdot \mathrm{day}^{-1} $ & \Sexpr{mysignif(parms$f_Ii,2)}\\

@@

$\alpha$ & Alga maximum exponential growth rate&$10^{-2} \cdot \mathrm{day} ^{-1} $& \Sexpr{mysignif(parms$alpha,2)} \\
$\gamma_n$ & Native host consumption rate&$10^{-7} \cdot \mathrm{L} \cdot \mathrm{individual}^{-1} \cdot \mathrm{day} ^{-1} $ & \Sexpr{mysignif(parms$gamman,2)}\\
$\gamma_i$ & Invasive host consumption rate&$10^{-6} \cdot \mathrm{L} \cdot \mathrm{individual}^{-1} \cdot \mathrm{day} ^{-1} $ & \Sexpr{mysignif(parms$gammai,2)}\\
$\mu$ & Alag refilling rate&$10^6 \mathrm{cell} \cdot \mathrm{L} ^{-1} \cdot \mathrm{day}^{-1}$ & 0.37\\
$\xi$ &Reduction in infected host reproduction and consumption& $\mathrm{Unitless}$ & \Sexpr{mysignif(parms$xi,2)}\\
$\beta_i$ & Spores produced per infected invasive species&$10^3 \cdot \mathrm{spore} \cdot \mathrm{individual}^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$betai,2)}\\
$\beta_n$ & Spores produced per infected native species&$10^3 \cdot \mathrm{spore} \cdot \mathrm{individual}^{-1} \cdot \mathrm{day}^{-1}$ & \Sexpr{mysignif(parms$betan,2)}\\
